Fractional Types

### 2013-03-10 rationals/rational_ext.tex
reorg

Negative types should correspond to sets with negative arity. But what
does that mean? Various approaches have been proposed -- Loeb uses the
notion of multiplicity functions to express negative sized sets
[CITE]. These notions have been extended to Euler characteristic of
categories by Leinster, Joyal and Schanuel -- negative values are
explained sometimes as equalities relating otherwise independent
objects -- Baez tells a story about ``island'' and positive and
``bridges connecting islands'' as negatives (since the effectively
reduce the number of disconnected elements.)

>>>>

Negative types should correspond to sets with negative arity. But what
does that mean? Various approaches have been proposed -- Loeb uses the
notion of multiplicity functions to express negative sized sets
[CITE]. These notions have been extended to Euler characteristic and
Homotopy cardinality of categories by Leinster, Joyal and Schanuel --
negative values are explained sometimes as equalities relating
otherwise independent objects -- Baez tells a story about ``island''
and positive and ``bridges connecting islands'' as negatives (since
the effectively reduce the number of disconnected elements.)

### 2013-03-10 rc2013/frac.tex
reorg

\newcommand{\roshan}[1]{} 
%% \newcommand{\roshan}[1]{\textsc{Roshan says:} 
%%   \textit{#1}
%% }

@@

\roshan{The above para is too technical in a weird way. Its unclear
  what it contributes. If one wants to make a point of computational
  witnesses of isomorphisms its easier to explain and cite in terms of
  Andreas Blass and Fiore -- and more generally by Barry Mazur. I say
  this also because it seems that HOTT is more concerned with
  topological structure of equalities, even though they tend to start
  their literature with some talk of intentional specifications.}

@@

\roshan{Again, HOTT seems jarring. Saying something about logic in
  this way however seems useful. Maybe one can use one of the examples
  of equivalence between proofs that Baez's Rosetta stone uses. }

Technically speaking, obtaining a higher-order extension requires the
construction of a \emph{closed category} from the underlying monoidal
category for~{{Pi}}. Although the general idea of such a construction is
well-understood, the details of adapting it to an actual programming language
are subtle.  Our main novel technical device to achieving the higher-order
extension is a \emph{fractional type} which represents \emph{negative
  information} and which is so named because of its duality with conventional
product types.  
%% By adding a formal (linear) dual to (type theoretic) products, we can
%% internalize the notion that ``taking an input'' corresponds
%% to negative information, which turns out to be the basis of first-class
%% functions.  
%% \roshan{The last sentence is terrible.}
The remainder of the paper reviews {{Pi}} and then introduces the syntax and
semantics of the extension with fractional types. We then study the
properties of the extended language and establish its expressiveness via
several constructions and examples that exploit its ability to model
higher-order computations.

%% AMR:
%% \roshan{It is unclear what the Id types refer to. We should provide
%%   some more context here}
%% I think the previous paragraph in the intro that talks about Martin-Lof type 
%% theory provides the context. Anything else??

@@

\roshan{In the above the, when we say that the terms of {{Pi}} are
  proofs that certain ``shapes of values'' are isomorphic, it is
  misleading. In {{langRev}}, the value {{left ()}} may or may not be
  isomorphic to the value {{right ()}} or {{left ()}} itself, or any
  other value of our choice. In particular the actual syntactic
  structure of the value has nothing to do with values being
  isomorphic. Values are just place-holders for the situations that
  their type represents. Their shapes have no intrinsic content. }

\roshan{Further, this is not to say that we may not be able to use the
  syntactic structure of a value for some purpose.  The prime use for
  such a thing might be, for example, to define a notion of distance
  between values -- such that values live in a metric space. Time and
  time again, we have felt the need for developing such an idea. Such
  a thing doesn't exist for the scope of this paper however. }

\roshan{The use of the word ``orthogonal'' is inappropriate in the
  footnote. Adding 0 truly affects how fractionals are developed,
  whereas when we say ``orthogonal'' we give the impression that it does
  not really matter. }

@@

\roshan{It is unclear what the usage ``first-class'' and ``effect of
  an isomorphism'' indicate. Maybe we should rephrase. }

@@

It is important to note that ``values'' and ``isomorphisms'' are completely
separate syntactic categories which do not intermix. The semantics of the
language come when these are made to interact at the ``top level'' via
\emph{application}: \roshan{I understand its intent, but this last sentence is unclear.}
%subcode{bnf} include main
% top level term, l ::= c v

@@

\roshan{Suggestion : This para is a good one to delete if we run out
  of space. This is largely general knowledge about universal algebra
  and maybe one useful remark about the categorical structure. However
  we have a whole paper about the categorical structure of Pi at RC
  itself. My point is this sort of this can be a remark inserted at
  other places in passing and it doesn't deserve a para on its own. It
  teaches the reader nothing new about Pi. }

@@

% \roshan{It is unclear how fractions help differentiate between inputs
%   and outputs. On first reading this seemed promising, but nothing
%   in the rest of the paper substantiates this point of view.}
% AMR: a function a -o b is modeled by (1/a,b) not (a,b) 
% Jacques: Exactly!  1/a is negative information is input, b is positive is
%    output.
%
% Roshan: I completely understand this and this is not my objection. This
% is not what the above sentence is saying however. Its talking about
% Pi not having a notion of input and output which is simply
% confusing.

% \roshan{I think we should arrive at relational programming as a means
%   of giving semantics to fractionals, rather than the other way
%   round. To show that relational programming is reversible, one does
%   not need to resort to fractionals. Everyone understand that
%   relations can be reversed and none of this setup with {{Pi}} is
%   required for that. Our key point should be to communicate that
%   fractionals have a clean semantics in the setting of relational
%   programming rather than the other way round.}
% AMR: relations are indeed only presented as a way to explain the semantics
% Roshan: it reads much better now.

@@

% \roshan{ In the rules above it is unclear as to which v is chosen in
%   the rule for eta. It is unclear that these rules are logically
%   reversible.  }

% \begin{proposition}[Logical Reversibility]
% \label{chx:prop:logrev-tracep}
% For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
% {{v2 : b2}} we have {{v1 ~~c~~ v2}} iff {{v2 ~~(sym c)~~ v1}}.
% \end{proposition}

% Some thoughts on a relational semantics. We define {{ a <-> b }}
% $\equiv$ {{a R b}}

% \noindent
% We now have primitives:
% %subcode{opsem} include main
% % swap*:b1*b2<->b2*b1 '= {[ ((v1, v2), (v2, v1)) ~|~ v1 in b1, v2 in b2 ]}
% % swap+:b1+b2<->b2+b1 '= {[ ((left v1), (right v1)) and ((right v2), (left v2)) 
% %                     && ~~~| v1 in b1, v2 in b2]}
% % assocl*:b1*(b2*b3)<->(b1*b2)*b3 '= {[ ((v1, (v2, v3)), ((v1, v2), v3)) 
% %                   & & ~~~|~ v1 in b1, v2 in b2, v3 in b3]}
% etc..

% \noindent
% closure combinators:
% %subcode{opsem} include main
% % sym c '= {[(v2, v1) ~|~ (v1, v2) in c ]}
% % c1 (;) c2 '= {[(v1, v3) ~|~ v2 in b2, (v1, v2) in c1, (v2, v3) in c2 ]}
% % c1 (*) c2 '= {[((v1, v2), (v3, v4)) ~|~ (v1, v3) in c1, (v2, v4) in c2 ]}
% etc..

% \noindent
% Now we have 
% %subcode{opsem} include main
% % eta* '= {[ ((), (v, 1/v)) ~|~ v in b ]}
% % eps* '= {[ ((v, 1/v), ()) ~|~ v in b ]}

% Now {{PiTF}} is logically reversible in that for all {{c:b1<->b2}} we
% have {{c^{dagger}: b2 <-> b1}} such that 
% %
% {{c^{dagger} = {[(v2, v1)~|~(v1,v2) in c ]} }}. This is
% straightforward from the definitions above.

% I think defining eval is an open problem.  Consider a few relations:

% \begin{enumerate}
% \item 
% {{empty = {[ ]} }}

% \item 
% {{full =  {[ R, G, B ]} * {[ R, G, B ]} }}

% \item 
% {{id  =  {[ (R, R), (G, G), (B, B) ]} }}

% \item 
% {{cycle = {[ (R, R), (R, G), (G, G), (G, B), (B, B), (B, R) ]} }}
% \end{enumerate}

% Can we define an {{eval}} that meaningfully distinguishes these? What
% can we say about {{(c (;) c^^^) v}} based on {{eval}}?  Here are
% several possible {{eval}} definitions:

% \begin{enumerate}
% \item {{ c v = v'}} where {{ v' in {[ v' ~|~ (v, v') in c ]} }}. The
%   eval maps {{b1 -> b2}}.

% \item {{ c v = {[ v' ~|~ (v, v') in c ]} }}. The eval maps {{b1 -> Pow(b2)}}.

% \item {{ c vs = {[ v' ~|~ (v, v') in c, v in vs ]} }}. where
%   {{vs:Pow(b1)}}. The eval maps {{Pow(b1) -> Pow(b2)}}.

% \end{enumerate}

% For the relations above, none of the suggested eval definitions give a
% logically reversible semantics.

% \roshan{If {{|b1|=|b2|}}, can we define all relations between {{b1}} and
% {{b2}}? If yes, then we have a model for {{PiTF}} where the objects
% are sets {{b}} and the morphisms are the relations between objects of
% equal size.}

% \roshan{ I suspect that all relations can be constructed, but we have
%   to show this. This is not obvious.}

% \roshan{In the judgments below it is misleading that only {{eta*}}
%   appears indexed by a type. To construct each family of relations,
%   the corresponding definition must be indexed by their types. In
%   other words, there is nothing special about {{eta*}}. Further, I am
%   unsure of how to format this so that the combinators are correctly
%   indexed by their types for both primitives and for composition
%   combinators. We could do the following, but it seems to be a
%   overkill.}

% %subcode{proof} include main
% %@ v1 in [[b1]]
% %@ v2 in [[b2]]
% %@ v3 in [[b3]]
% %@@ (v1, (v2, v3)) ~assocl*~ ((v1, v2), v3)

@@

\roshan{Either we should have explain the SAT solver well or we
  shouldn't have at all. There are some original ideas in the
  construction of that solver and the Oxford QF group for instance
  were very interested in its details. This presentation neither
  leaves the idea prime for future publication, and neither does it
  explain it well.}

@@

\roshan{Again, if we have lost zero, we cannot talk about vector
  spaces with the zero vector. We should alternately included it in
  our presentation to start with. }

@@

\roshan{The above para should simply be deleted.}

@@

\roshan{Again, the above para is just lose speculation. Further, the
  fact that we say that its cartesian closed instead of monoidal
  closed should *immediately* tell us that its the wrong model.}

>>>>

@@

@@

Technically speaking, obtaining a higher-order extension requires the
construction of a \emph{closed category} from the underlying monoidal
category for~{{Pi}}. Although the general idea of such a construction is
well-understood, the details of adapting it to an actual programming language
are subtle.  Our main novel technical device to achieving the higher-order
extension is a \emph{fractional type} which represents \emph{negative
  information} and which is so named because of its duality with conventional
product types.  
%% By adding a formal (linear) dual to (type theoretic) products, we can
%% internalize the notion that ``taking an input'' corresponds
%% to negative information, which turns out to be the basis of first-class
%% functions.  
The remainder of the paper reviews {{Pi}} and then introduces the syntax and
semantics of the extension with fractional types. We then study the
properties of the extended language and establish its expressiveness via
several constructions and examples that exploit its ability to model
higher-order computations.

%% AMR:
%% I think the previous paragraph in the intro that talks about Martin-Lof type 
%% theory provides the context. Anything else??

@@

@@

@@

It is important to note that ``values'' and ``isomorphisms'' are completely
separate syntactic categories which do not intermix. The semantics of the
language come when these are made to interact at the ``top level'' via
\emph{application}: 
%subcode{bnf} include main
% top level term, l ::= c v

@@

@@

% AMR: a function a -o b is modeled by (1/a,b) not (a,b) 
% Jacques: Exactly!  1/a is negative information is input, b is positive is
%    output.

% AMR: relations are indeed only presented as a way to explain the semantics

@@

@@

@@

@@

@@
